Bird beak's elaboration and innovation

### 2023-04-12 Analysis/14-Elab_innov_through_time.Rmd
Create 14-Elab_innov_through_time.Rmd

This script reproduces a figure not present or used in the manuscript or the supplementary materials.

@@

# All birds analysis

@@

Extract the median elaboration and innovation from the results for the mega level (projection on the phylogeny) and the macro level (projection on each group).

```{r, eval = TRUE, echo = TRUE}
## Calculate the correlations for the projections onto the phylo axis
super_phylo_cor <- get.correlations.plot(results$super_results_phylo)

## Calculate the correlations for the projections onto the group' axes
super_group_cor <- get.correlations.plot(results$super_results_within, exclude.phylo = TRUE)
```

## Get the correlation fable

y axis = levels
column 1 = correlation of e/i projected on the mega level
column 2 = correlation of e/i projected on the macro level
stars for significant correlation between means.

```{r, echo = TRUE, eval = TRUE}
## Get all the correlations for all the clades
cor_mega_level <- get.correlations.scores(results$super_results_phylo)
cor_macro_level <- get.correlations.scores(results$super_results_within)

## Get the colours
load("../Data/Processed/tip_colours_super_orders.rda")
super_orders_col <- c(tip_colours_super_orders, "phylogeny" = "grey")

## Plot the correlations
pdf("../Manuscript/Figures/correlations_fable.pdf")
par(mfrow = c(1,2))
par(mar = c(5, 8, 4, 0) + 0.1, bty = "n") # c(bottom, left, top, right)
plot.cor.scores(cor_mega_level, col = super_orders_col,
                y.axis = TRUE, plot.phylo = TRUE,
                main = "Elaboration and innovation\nat the mega level") 
par(mar = c(5, 2, 4, 6) + 0.1, bty = "n") # c(bottom, left, top, right)
plot.cor.scores(cor_macro_level, col = super_orders_col,
                y.axis = FALSE, plot.phylo = FALSE,
                main = "Elaboration and innovation\nat the macro level")
dev.off()
```

## Ploting the figure

The following snippets plots the entire figure minus the panel IDs:
 * Panel A: the overal density figure
 * Panel B: the correlation plots for the projection on the mega level
 * Panel C: the correlation plots for the projection on the macro level

Density plot of the correlation of the median elaboration and median exploration for all species.

```{r, echo = TRUE, eval = TRUE}
## Set the pdf parameters
pdf("../Manuscript/Figures/correlations.pdf", height = 12, width = 8)

## Set up the plotting layout
layout_matrix <- cbind(matrix(1,2,2), matrix(0,2,2))
layout_matrix <- rbind(rbind(layout_matrix, matrix(1:8+1, 2, 4, byrow = TRUE)), matrix(9:16+1, 2, 4,  byrow = TRUE))
plot_layout <- layout(layout_matrix)
# layout.show(plot_layout)

## Setting the colour pallette for the subsequent plots
colour.palette <- gg.color.hue

## --------
## PANEL A
## --------

## Plotting the densities (with the quadrants) in the first panel
par(mar = c(5, 4, 4, 2) + 0.1)
plot.densities(super_phylo_cor$phylogeny, with.quadrant = TRUE)

## --------
## PANEL B
## --------

## Getting the subsets for the colouring
col_subs <- c(split(shapespace$level2, f = shapespace$level1)[-1], "phylogeny" = list(shapespace$level1))

## Set the vector of legend positions
legends_pos <- c("topleft", "topright", "topright", "topright",
                 "topleft", "topright", "topleft", "topright")
## Create the vector for counting the legend numbers
legend_counter <- 1

## Plotting each individual correlation plots
for(i in 1:8) {
    par(mar = c(2,2.5,2,1.5))
    plot.correlations(super_phylo_cor[[i]], col.sub = col_subs[[i]], legend.pos = legends_pos[i], ID = i, pt.cex = 0.5, legend.cex = 2/3, legend_counter = legend_counter, contour = NULL, with.quadrant = TRUE)
}

## --------
## PANEL C
## --------

## Set the vector of legend positions
legends_pos <- c("bottomright", "topright", "topleft", "topleft",
                 "topleft", "topleft", "topright", "bottomright")
## Resetting the legend_counter
legend_counter <- 1

## Plotting each individual correlation plots
for(i in 1:8) {
    par(mar = c(2,2.5,2,1.5))
    plot.correlations(super_group_cor[[i]], col.sub = col_subs[[i]], legend.pos = legends_pos[i], ID = i, pt.cex = 0.5, legend.cex = 2/3, legend_counter = legend_counter, with.quadrant = FALSE, contour = NULL)
}
dev.off()
```

Extract the median elaboration and innovation from the results for the macro level (projection on the passeriformes) and the macro level-1 (projection on each group).

# Passeriformes analysis

```{r, eval = TRUE, echo = TRUE}
## Load the data
load(file = "../Data/Processed/shapespace_passeriformes_two_levels.rda")
## Select the data
shapespace <- shapespace_passeriformes_two_levels[[1]]$space
tree <- shapespace_passeriformes_two_levels[[1]]$consensus_tree
## Clean garbage
rm(shapespace_passeriformes_two_levels)

## Load the covar
load(file = "../Data/Processed/shapespace_passeriformes.MCMCglmm")
shapespace_passeriformes <- shapespace_passeriformes[[1]]

## Get the levels
phylo_lvl <- which(MCMCglmm.levels(shapespace_passeriformes) == "animal")
subor_lvl <- grep("level1_", MCMCglmm.levels(shapespace_passeriformes))
famil_lvl <- grep("level2_", MCMCglmm.levels(shapespace_passeriformes))

## Creating the subsets
suborder_shapespace <- MCMCglmm.subsets(#n = 1000, # Just run it on 1k matrices for now
      data          = shapespace,
      posteriors    = shapespace_passeriformes,
      group         = MCMCglmm.levels(shapespace_passeriformes)[c(subor_lvl, phylo_lvl)],
      rename.groups = c(levels(shapespace$level1)[-1], "phylogeny")) #[-1] is because the first level is ""

family_shapespace <- MCMCglmm.subsets(#n = 1000, # Just run it on 1k matrices for now
      data          = shapespace,
      posteriors    = shapespace_passeriformes,
      group         = MCMCglmm.levels(shapespace_passeriformes)[c(famil_lvl, phylo_lvl)],
      rename.groups = c(levels(shapespace$level2)[-1], "phylogeny")) #[-1] is because the first level is ""
## Clean garbage
rm(shapespace_passeriformes)

## Loading the results
load(file = "../Data/Processed/shapespace_passeriformes_suborder_family_results_list.rda")
results <- shapespace_passeriformes_suborder_family_results_list
rm(shapespace_passeriformes_suborder_family_results_list)
```

```{r, eval = TRUE, echo = FALSE}
## Calculate the correlations for the projections onto the phylo axis
suborder_phylo_cor <- get.correlations.plot(results$suborder_results_phylo)

## Calculate the correlations for the projections onto the group' axes
suborder_group_cor <- get.correlations.plot(results$suborder_results_within, exclude.phylo = TRUE)
```

## Ploting the figure

The following snippets plots the entire figure minus the panel IDs:
 * Panel A: the overal density figure
 * Panel B: the correlation plots for the projection on the mega level
 * Panel C: the correlation plots for the projection on the macro level

Density plot of the correlation of the median elaboration and median exploration for all species.

```{r, echo = TRUE, eval = TRUE}
## Set the pdf parameters
pdf("../Manuscript/Figures/correlations_passeriformes.pdf", height = 12, width = 6)

## Set up the plotting layout
layout_matrix <- cbind(matrix(1,2,2), matrix(0,2,1))
layout_matrix <- rbind(rbind(layout_matrix, matrix(1:6+1, 2, 3, byrow = TRUE)), matrix(7:12+1, 2, 3,  byrow = TRUE))
plot_layout <- layout(layout_matrix)
# layout.show(plot_layout)

## Setting the colour pallette for the subsequent plots
colour.palette <- gg.color.hue

## --------
## PANEL A
## --------

## Plotting the densities (with the quadrants) in the first panel
par(mar = c(5, 4, 4, 2) + 0.1)
plot.densities(suborder_phylo_cor$phylogeny, with.quadrant = TRUE)

## --------
## PANEL B
## --------

## Getting the subsets for the colouring
col_subs <- c(split(shapespace$level2, f = shapespace$level1)[-1], "phylogeny" = list(shapespace$level1))

## Set the vector of legend positions
legends_pos <- c("topright", "topright", "topright", "topright", "topright", "topright")
## Create the vector for counting the legend numbers
legend_counter <- 1

## Plotting each individual correlation plots
for(i in 1:6) {
    par(mar = c(2,2.5,2,1.5))
    plot.correlations(suborder_phylo_cor[[i]], col.sub = col_subs[[i]], legend.pos = legends_pos[i], ID = i, pt.cex = 0.5, legend.cex = 2/3, legend_counter = legend_counter, contour = NULL, with.quadrant = TRUE)
}

## --------
## PANEL C
## --------

## Set the vector of legend positions
legends_pos <- c("topright", "topright", "topright", "topright", "topright", "topright")

## Resetting the legend_counter
legend_counter <- 1

## Plotting each individual correlation plots
for(i in 1:6) {
    par(mar = c(2,2.5,2,1.5))
    plot.correlations(suborder_group_cor[[i]], col.sub = col_subs[[i]], legend.pos = legends_pos[i], ID = i, pt.cex = 0.5, legend.cex = 2/3, legend_counter = legend_counter, with.quadrant = FALSE, contour = NULL)
}
dev.off()
```

Legend numbers are:
1 - others (groups with < 15 species)
2 - Corvoidea
3 - Malaconotoidea
4 - Orioloidea
5 - Meliphagoidea
6 - Bombycilloidea
7 - Muscicapoidea
8 - Emberizoidea
9 - Motacillidae
10 - Nectariniidae 
11 - Passeridae
12 - PloceidaeEstrildidae
13 - Eurylaimides
14 - Furnariida
15 - Tyrannida 
16 - Aegithaloidea
17 - Cisticolidae
18 - Fringillidae
19 - Hirundinidae
20 - Locustelloidea
21 - Paridae
22 - Pycnonotidae
23 - Sylvioidea

>>>>

@@

@@

# Checking elaboration and innovation values through time (clades)

```{r, eval = TRUE, echo = TRUE}
## Get all the statistics
super_orders_elaboration <- summary(results$super_results_between$position)
super_orders_innovation <- summary(results$super_results_between$distance)
orders_elaboration <- summary(results$order_results_between$position)
orders_innovation <- summary(results$order_results_between$distance)

## Get all the tree ages
load("../Data/Processed/ploting_data.rda")
load("../Data/Processed/clade_info.rda")

get.clade.age <- function(name, info, tree) {
  return(list(age = max(dispRity::tree.age(keep.tip(tree, info[[name]]$species))$ages), col = info[[name]]$col))
}

super_age_info <- t(sapply(unlist(strsplit(super_orders_elaboration$subsets, split = ":phylogeny")), get.clade.age, info = clade_info$super_orders, tree = ploting_data$consensus_tree))
order_age_info <- t(sapply(unlist(strsplit(orders_elaboration$subsets, split = ":phylogeny")), get.clade.age, info = clade_info$orders, tree = ploting_data$consensus_tree))

## Table that for easy plotting
super_orders_elaboration <- cbind(super_orders_elaboration, super_age_info, cex = 2)
super_orders_innovation <- cbind(super_orders_innovation, super_age_info, cex = 2)
orders_elaboration <- cbind(orders_elaboration, order_age_info, cex = 1)
orders_innovation <- cbind(orders_innovation, order_age_info, cex = 1)
elaboration_plot_data <- rbind(super_orders_elaboration, orders_elaboration)
innovation_plot_data <- rbind(super_orders_innovation, orders_innovation)

## Empty plot
plot.age.data <- function(data, ylab, xlab) {
  plot(NULL, xlab = xlab, ylab = ylab,
       xlim = rev(range(pretty(range(data$age)))),
       ylim = range(pretty(range(data[,c(4, 8)]))))
  ## Adding the lines for each clade
  for(i in 1:nrow(data)) {
    lines(x = rep(data[i, "age"][[1]], 2), y = data[i, c("2.5%", "97.5%")], lty = 2, col = data[i, "col"][[1]])
    lines(x = rep(data[i, "age"][[1]], 2), y = data[i, c("25%", "75%")], lty = 1, lwd = 2, col = data[i, "col"][[1]])
    points(x = data[i, "age"][[1]], y = data[i, "obs.median"], cex = data[i, "cex"], col = data[i, "col"][[1]], pch = 19)
  }
}
pdf(height = 14, width = 7, file = "../Manuscript/Figures/Elaboration_and_innovation_through_time.pdf")
par(mfrow = c(2,1))
plot.age.data(elaboration_plot_data, ylab = "Elaboration", xlab = "")
legend("topleft", pch = 19, col = "grey", legend = c("super orders", "orders"), pt.cex = c(2, 1))
plot.age.data(innovation_plot_data, ylab = "Innovation", xlab = "Time (Mya)")
dev.off()
```
